
---
output: 
  html_document:
  pdf_document: default
  word_document: default
title: "Final Project - Building Interactive Dashboard"
---

------------------------------------------------------------------------

**Submission Guideline:** Create one html file from Rmarkdown that answer all the following questions. **Each question should has it own Rmd file and a corresponding link**. Include the link of the html file to your Github webpage and submit the link to the html file to Canvas.

------------------------------------------------------------------------

### Question 1.

-   Install the flexdashboard package using `install.packages("flexdashboard")`.

install.packages("flexdashboard")

-   Create an empty dashboard. In Rstudio, File -\> New File -\> R Markdown -\> From Template -\> Flex Dashboard. Knit the rmarkdown file and add the html link to this report.



### Question 2.

Use this template flex dashboard ([dashboard](https://bryantstats.github.io/math421/final/flex_template.html), [Rmd](https://bryantstats.github.io/math421/final/flex_template.Rmd)) to create a flex dashboard that
```{r}
# Use this template flex dashboard to create a flex dashboard that
# - Has a side bar
# - Has at least three main Tabs
# - Has at least four plots
# - Shows a table data that can be searched
# Link of the dashboard:
```
[Final Q1+Q2.html](https://github.com/esmith28-1/Math421/blob/main/Final-Q1-Q2.html)

### Question 3.

```{r}
## We want users to be able to interact with our dashboard. This can be done using `shiny`. A flex dashboard using `shiny` requires a domain to host it. We will use shinyapps.io, which can host five dashboard for free. If we want to host more than five dashboards, we can create multiple accounts at shinyapps.io.
## Follow these steps to create a shiny flex dashboard then publish it. Include the link of the dashboard here.
```

Dashboard Link: [Question 3 Dashboard](https://esmith28.shinyapps.io/Final_Q3/)

### Question 4 - selectInput

```{r}
## Write a shiny flex dashboard that uses `selectInput` and works with a dataset different from the titanic dataset.
## Note: In the above dashboard, click to Source Code to show the codes.
## Include the link of your dashboard here.
```

Include the link of your dashboard here.
[Question 4 Dashboard](https://esmith28.shinyapps.io/Final_Q4/)


### Question 5 - checkboxGroupInput
```{r}
## Write a shiny flex dashboard that uses `checkboxGroupInput` and works with a dataset different from the titanic dataset.
```

Include the link of your dashboard here.
[Question 5 Dashboard](https://esmith28.shinyapps.io/Final_Q5/)

### Question 6 - sliderInput
```{r}
## Write a shiny flex dashboard that uses `sliderInput` and works with a dataset different from the titanic dataset.
```

Include the link of your dashboard here.
[Question 6 Dashboard](https://esmith28.shinyapps.io/Final_Q6/)

### Question 7 - radioButtons
```{r}
## Write a shiny flex dashboard that uses `radioButtons` and works with a dataset different from the titanic dataset.
```

Include the link of your dashboard here.
[Question 7 Dashboard](https://esmith28.shinyapps.io/Final_Q7/)

------------------------------------------------------------------------

**Notice: You can only run maximum 5 apps on shinyapps.io. If you are running out of apps, you are recommended to create another shiny account and use multiple accounts.**

------------------------------------------------------------------------

### Question 8 - dateRangeInput
```{r}
## Write a shiny flex dashboard that uses `dateRangeInput` and works with a dataset different from the titanic dataset.
```

Include the link of your dashboard here.
[Question 8 Dashboard](https://esmith27.shinyapps.io/Final_Q8/)

### Question 9 - Combination of Inputs
```{r}
## We can combine different input methods into one dashboard. Write a shiny flex dashboard and works with a dataset different from the titanic dataset. The dashboard should use at least three different input methods.
```

Include the link of your dashboard here.
[Question 9 Dashboard](https://esmith27.shinyapps.io/Final_Q9/)

------------------------------------------------------------------------

### Question 10
```{r}
## Write a dashboard that
##-  Works with a dataset that have not been used in class
##-  Has a social media click
##-  Has a source code click
##-  Has a side bar
##-  Has five input methods
##-  Has at least three main tabs
##-  Has columns tabs
##-  Has at least five meaningful plots
##-  Shows a table data that can be searched
##-  Has at least one text paragraph
##-  Includes at least one image
```

Include the link of your dashboard here.

[Final Dashboard](https://esmith27.shinyapps.io/Final_Q10/)

------------------------------------------------------------------------

### Question 11 - Presentation

Present your dashboard in Question 10 on the final exam day.

-   Present the features of your dashboard

-   Discuss how you organize the codes of the dashboard and the final project.

-   Discuss any error or challenge you had to overcome

-   Give an advice for anyone who is about to do this project

------------------------------------------------------------------------

### References

<https://rstudio.github.io/flexdashboard/articles/examples.html>

<https://bookdown.org/yihui/rmarkdown/dashboards.html>

<https://epirhandbook.com/en/dashboards-with-r-markdown.html>

Check out Shiny official tutorial: <https://shiny.rstudio.com/tutorial/>

Check out Shiny Galary: <https://shiny.rstudio.com/gallery/>

Pairs of render and output: <https://bookdown.org/weicheng/shinyTutorial/images/outputs.png>

Check out some shiny layouts here: <https://shiny.rstudio.com/articles/layout-guide.html>

Check out shiny dynamic UI here

<https://shiny.rstudio.com/gallery/dynamic-ui.html>

<https://shiny.rstudio.com/articles/dynamic-ui.html>

------------------------------------------------------------------------
