---
title: "Midterm Presentation"
format: revealjs
theme: simple
output-file: "Midterm_Presentation.html"
---

## Rhode Island Department of Health Data Presentation

by Emerson Smith

------------------------------------------------------------------------

## Distribution of Age

```{r setup, include=FALSE, cache=TRUE}
getwd()
load("df_loaded.RData")
load("df2_loaded.RData")
load("df3_loaded.RData")
```

```{r}
library(lubridate)
library(tidyverse)
library(knitr)
plot1 <- ggplot(df, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Age",
       x = "Age",
       y = "Count",
       caption = "Histogram showing the distribution of patient ages") +
  theme_minimal()
print(plot1)
```

------------------------------------------------------------------------

## Total Charges by Month of Admission

```{r}
plot2 <- ggplot(df, aes(x = factor(moa), y = tot)) +
  geom_boxplot(fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "Total Charges by Month of Admission",
       x = "Month of Admission",
       y = "Total Charges",
       caption = "Boxplot showing total charges across different months of admission") +
  theme_minimal()
print(plot2)
```

------------------------------------------------------------------------

## Distribution of Length of Stay

```{r}
plot3 <- ggplot(df, aes(x = los)) +
  geom_histogram(binwidth = 1, fill = "pink", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Length of Stay",
       x = "Length of Stay (days)",
       y = "Count",
       caption = "Histogram showing the distribution of length of stay") +
  theme_minimal()
print(plot3)
```

------------------------------------------------------------------------

## Average Length of Stay by Season

```{r}
plot4 <- ggplot(df, aes(x = factor(season), y = los)) +
  geom_bar(stat = "summary", fun = "mean", fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Average Length of Stay by Season",
       x = "Season",
       y = "Average Length of Stay (days)",
       caption = "Bar plot showing average length of stay across different seasons") +
  theme_minimal()
print(plot4)
```

------------------------------------------------------------------------

## Total Charges by Provider

```{r}
plot5 <- ggplot(df, aes(x = reorder(provider, -tot), y = tot)) +
  geom_bar(stat = "summary", fun = "sum", fill = "purple", color = "black", alpha = 0.7) +
  labs(title = "Total Charges by Provider",
       x = "Provider",
       y = "Total Charges",
       caption = "Bar plot showing total charges by each provider") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
print(plot5)
```

------------------------------------------------------------------------

## Length of Stay vs. Total Charges

```{r}
plot6 <- ggplot(df, aes(x = los, y = tot)) +
  geom_point(alpha = 0.5, color = "red") +
  labs(title = "Length of Stay vs Total Charges",
       x = "Length of Stay (days)",
       y = "Total Charges",
       caption = "Scatter plot showing the relationship between length of stay and total charges") +
  theme_minimal()
print(plot6)
```

------------------------------------------------------------------------

## Average Total Charges by Age Group

```{r}
plot7 <- ggplot(df, aes(x = cut(age, breaks = seq(0, 100, by = 10)), y = tot)) +
  geom_bar(stat = "summary", fun = "mean", fill = "cyan", color = "black", alpha = 0.7) +
  labs(title = "Average Total Charges by Age Group",
       x = "Age Group",
       y = "Average Total Charges",
       caption = "Bar plot showing average total charges across different age groups") +
  theme_minimal()
print(plot7)
```

------------------------------------------------------------------------

## Distribution of Total Charges

```{r}
plot8 <- ggplot(df, aes(x = tot)) +
  geom_histogram(binwidth = 1000, fill = "yellow", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Total Charges",
       x = "Total Charges",
       y = "Count",
       caption = "Histogram showing the distribution of total charges") +
  theme_minimal()
print(plot8)
```

------------------------------------------------------------------------

## Average Total Charges by Season

```{r}
plot9 <- ggplot(df, aes(x = factor(season), y = tot)) +
  geom_bar(stat = "summary", fun = "mean", fill = "lightgreen", color = "black", alpha = 0.7) +
  labs(title = "Average Total Charges by Season",
       x = "Season",
       y = "Average Total Charges",
       caption = "Bar plot showing average total charges across different seasons") +
  theme_minimal()
print(plot9)
```

------------------------------------------------------------------------

## Average Age by Month of Admission

```{r}
plot10 <- ggplot(df, aes(x = factor(moa), y = age)) +
  geom_bar(stat = "summary", fun = "mean", fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Average Age by Month of Admission",
       x = "Month of Admission",
       y = "Average Age",
       caption = "Bar plot showing average age across different months of admission") +
  theme_minimal()
print(plot10)
```

------------------------------------------------------------------------

## Total Charges Over Months (Animated)

```{r, warning=FALSE}
library(ggplot2)
library(gganimate)
library(dplyr)
plot11 <- ggplot(df, aes(x = factor(moa), y = tot, group = 1)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  labs(title = "Total Charges Over Months",
       subtitle = "Month: {frame_along}",
       x = "Month of Admission",
       y = "Total Charges",
       caption = "Line plot showing total charges over different months of admission") +
  theme_minimal() +
  transition_reveal(moa)
animate(plot11, nframes = 100, fps = 10)
```

------------------------------------------------------------------------

## Target 1 Training Data rpart (Model 1.1)

```{r}
library(rpart)
library(rpart.plot)
library(caret)
dt_model <- rpart(target ~ ., 
            data = df2_train,
            control=rpart.control(maxdepth=2))
rpart.plot(dt_model) 
if (!is.null(dt_model$splits) && nrow(dt_model$splits) > 0) {
  varImp <- varImp(dt_model)
  print(varImp)
} else {
  cat("Tree has no splits — variable importance not available.\n")
}
```

------------------------------------------------------------------------

## Caret Model for Target 1 (Model 1.2 )

```{r}
library(caret)
## Model 1.2
tuneGrid <- expand.grid(maxdepth = 2:10)
trControl <- trainControl(method = "cv",
                         number = 5)
dt_model2 <- train(
  target ~ ., 
  data = df2_train,
  method = "rpart2", 
  trControl = trControl,
  tuneGrid = tuneGrid,
  na.action = na.omit
)
plot(dt_model2)
```

------------------------------------------------------------------------

## Caret Model for Target 1 (Model 1.3)

```{r}
knn_model <- train(
  target ~ .,
  data = df2_train,
  method = "knn",
  trControl = trControl,
  tuneLength = 10,
  na.action = na.omit
)
knn_model
plot(knn_model)
```

------------------------------------------------------------------------

## Model Comparison for Target 1

```{r}
results <- resamples(list(DecisionTree = dt_model2, KNN = knn_model))
bwplot(results)
```

------------------------------------------------------------------------

## Final Selection for Target 1

```{r}
library(caret)
prediction <-predict(dt_model2, newdata = df2_test)
confusionmatrix <-confusionMatrix(data = prediction, reference = df2_test$target, positive = "high")
confusionmatrix$overall[1]
```

------------------------------------------------------------------------

## New Target Variable (in df3_loaded.RData)

df3 \<- df3 %\>%

mutate(target = case_when(

sex == 1 \~ "Male",

sex == 2 \~ "Female",

TRUE \~ NA_character\_

))

filter(!is.na(target))

```{r}
table(df3$target, useNA = "ifany")
df3 <- df3 %>%
  select(age, raceethn, provider, moa, mod, admtype, campus, total, los, target)
```

------------------------------------------------------------------------

## Testing and Training Data

```{r}
library(tidyverse)
library(caret)
library(rpart)
df3 <- df3 %>% 
  mutate(target = as.factor(target),
         raceethn = as.factor(raceethn),
         age = as.numeric(age), 
         provider = as.factor(provider), 
         campus = as.factor(campus),
         admtype = as.factor(admtype), 
         moa = as.factor(moa), 
         mod = as.factor(mod),
         total = as.numeric(total))
splitIndex <- createDataPartition(df3$target, p = .10, 
                                  list = FALSE)
df3_train <- df3[ splitIndex,]
df3_test <- df3[-splitIndex,]
```

------------------------------------------------------------------------

## Target 2 Training Data rpart (Model 2.1)

```{r}
dt_model2 <- rpart(target ~ ., 
            data = df3_train,
            control=rpart.control(maxdepth=4))
library(rpart.plot)
library(caret)
rpart.plot(dt_model2)
```

------------------------------------------------------------------------

## Caret Model for Target 2 (Model 2.2)

```{r}
tuneGrid2 <- expand.grid(maxdepth = 2:10)
trControl2 <- trainControl(method = "cv",
                         number = 5)
dt_model3 <- train(target~., data=df3_train, 
                                method = "rpart2", 
                                trControl = trControl2,
                                tuneGrid = tuneGrid2)
plot(dt_model3)
```

------------------------------------------------------------------------

## Caret Model for Target 2 (Model 2.3)

```{r}
knn_model2 <- train(
  target ~ .,
  data = df3_train,
  method = "knn",
  trControl = trControl2,
  tuneLength = 10
)
knn_model2
plot(knn_model2)
```

------------------------------------------------------------------------

## Comparison Models for Target 2

```{r}
results2 <- resamples(list(DecisionTree = dt_model3, KNN = knn_model2))
bwplot(results2)
```

------------------------------------------------------------------------

## Final Model Selection for Target 2

```{r}
prediction2 <-predict(dt_model3, newdata = df3_test)
confusionmatrix2 <- confusionMatrix(data = prediction2, reference = df3_test$target, positive = "Male")
confusionmatrix2$overall[1]
```
